\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}

\vspace{0.5cm}

This project presents Threat Sentinel, an integrated cybersecurity threat analysis and visualization platform designed to address the increasing complexity of modern cybersecurity challenges. Organizations face difficulties in monitoring, analyzing, and responding to security threats across distributed network infrastructures. Traditional security tools often operate in isolation, lacking comprehensive integration and real-time visualization capabilities.

Threat Sentinel provides a unified solution that combines network scanning, vulnerability assessment, and threat intelligence into a single, intuitive platform. The system employs a modern web-based architecture with a React TypeScript frontend and a Python FastAPI backend, enabling real-time monitoring and analysis. Key features include automated host discovery, port scanning, service enumeration, vulnerability detection, and interactive data visualization through customizable dashboards.

The platform successfully demonstrates the effectiveness of integrating multiple security tools (Nmap, Masscan, Nikto) with modern web technologies to create an accessible and powerful threat analysis system. Through comprehensive testing and validation, the system proves capable of efficiently scanning network infrastructure, identifying potential security vulnerabilities, and presenting findings through clear, actionable visualizations. This project contributes to the field by demonstrating how modern web technologies can enhance traditional cybersecurity tools, making advanced security analysis accessible to organizations of varying technical capabilities.
