\appendix

\chapter{User Questionnaire}
\label{app:questionnaire}

This appendix contains the questionnaire used for gathering user feedback during the usability testing phase of the project.

\section{Background Information}

\begin{enumerate}
    \item What is your current role or position?
    \begin{itemize}
        \item[$\square$] Security Professional
        \item[$\square$] System Administrator
        \item[$\square$] Network Engineer
        \item[$\square$] Developer
        \item[$\square$] Student
        \item[$\square$] Other: \underline{\hspace{5cm}}
    \end{itemize}
    
    \item How many years of experience do you have in IT/cybersecurity?
    \begin{itemize}
        \item[$\square$] Less than 1 year
        \item[$\square$] 1-3 years
        \item[$\square$] 3-5 years
        \item[$\square$] 5-10 years
        \item[$\square$] More than 10 years
    \end{itemize}
    
    \item Have you previously used network scanning tools (Nmap, Masscan, etc.)?
    \begin{itemize}
        \item[$\square$] Yes, extensively
        \item[$\square$] Yes, occasionally
        \item[$\square$] Yes, but rarely
        \item[$\square$] No, never
    \end{itemize}
\end{enumerate}

\section{System Usability}

Please rate the following statements on a scale of 1-5:\\
(1 = Strongly Disagree, 2 = Disagree, 3 = Neutral, 4 = Agree, 5 = Strongly Agree)

\begin{enumerate}
    \item The system interface is intuitive and easy to navigate.
    \begin{itemize}
        \item[$\square$] 1 \quad [$\square$] 2 \quad [$\square$] 3 \quad [$\square$] 4 \quad [$\square$] 5
    \end{itemize}
    
    \item I could perform basic scans without requiring assistance or documentation.
    \begin{itemize}
        \item[$\square$] 1 \quad [$\square$] 2 \quad [$\square$] 3 \quad [$\square$] 4 \quad [$\square$] 5
    \end{itemize}
    
    \item The visualizations effectively communicate security information.
    \begin{itemize}
        \item[$\square$] 1 \quad [$\square$] 2 \quad [$\square$] 3 \quad [$\square$] 4 \quad [$\square$] 5
    \end{itemize}
    
    \item The system provides adequate feedback during scan operations.
    \begin{itemize}
        \item[$\square$] 1 \quad [$\square$] 2 \quad [$\square$] 3 \quad [$\square$] 4 \quad [$\square$] 5
    \end{itemize}
    
    \item Error messages are clear and help me understand what went wrong.
    \begin{itemize}
        \item[$\square$] 1 \quad [$\square$] 2 \quad [$\square$] 3 \quad [$\square$] 4 \quad [$\square$] 5
    \end{itemize}
    
    \item The dashboard provides a useful overview of security status.
    \begin{itemize}
        \item[$\square$] 1 \quad [$\square$] 2 \quad [$\square$] 3 \quad [$\square$] 4 \quad [$\square$] 5
    \end{itemize}
    
    \item I would recommend this system to colleagues who need security scanning capabilities.
    \begin{itemize}
        \item[$\square$] 1 \quad [$\square$] 2 \quad [$\square$] 3 \quad [$\square$] 4 \quad [$\square$] 5
    \end{itemize}
\end{enumerate}

\section{Functionality Assessment}

\begin{enumerate}
    \item Which scanning features did you use? (Check all that apply)
    \begin{itemize}
        \item[$\square$] Host scanning
        \item[$\square$] Port scanning
        \item[$\square$] Web vulnerability scanning
        \item[$\square$] Dashboard viewing
        \item[$\square$] Result export
    \end{itemize}
    
    \item Did you encounter any errors or unexpected behavior?
    \begin{itemize}
        \item[$\square$] Yes \quad [$\square$] No
    \end{itemize}
    
    If yes, please describe: \underline{\hspace{10cm}}
    
    \vspace{0.5cm}
    
    \item Were scan results accurate based on your knowledge of the target systems?
    \begin{itemize}
        \item[$\square$] Yes, completely accurate
        \item[$\square$] Mostly accurate
        \item[$\square$] Partially accurate
        \item[$\square$] Not accurate
        \item[$\square$] Unable to verify
    \end{itemize}
\end{enumerate}

\section{Performance}

\begin{enumerate}
    \item How would you rate the system's performance?
    \begin{itemize}
        \item[$\square$] Excellent
        \item[$\square$] Good
        \item[$\square$] Acceptable
        \item[$\square$] Poor
        \item[$\square$] Very Poor
    \end{itemize}
    
    \item Were there any noticeable delays or slow operations? If yes, where?
    
    \vspace{1cm}
    \underline{\hspace{13cm}}
    
    \vspace{0.5cm}
    \underline{\hspace{13cm}}
\end{enumerate}

\section{Open Feedback}

\begin{enumerate}
    \item What features did you find most valuable?
    
    \vspace{1.5cm}
    \underline{\hspace{13cm}}
    
    \vspace{0.5cm}
    \underline{\hspace{13cm}}
    
    \item What improvements would you suggest?
    
    \vspace{1.5cm}
    \underline{\hspace{13cm}}
    
    \vspace{0.5cm}
    \underline{\hspace{13cm}}
    
    \item Are there any features you feel are missing?
    
    \vspace{1.5cm}
    \underline{\hspace{13cm}}
    
    \vspace{0.5cm}
    \underline{\hspace{13cm}}
    
    \item Additional comments:
    
    \vspace{2cm}
    \underline{\hspace{13cm}}
    
    \vspace{0.5cm}
    \underline{\hspace{13cm}}
    
    \vspace{0.5cm}
    \underline{\hspace{13cm}}
\end{enumerate}

\vspace{1cm}

\textbf{Thank you for your participation!}

\clearpage

\chapter{Interview Questions}
\label{app:interview}

For more detailed feedback, the following semi-structured interview questions were used with selected participants:

\section{General Experience}

\begin{enumerate}
    \item Can you walk me through your first impression of the system?
    \item What was your primary goal when using Threat Sentinel?
    \item How does this system compare to other security tools you've used?
\end{enumerate}

\section{Specific Features}

\begin{enumerate}
    \item How intuitive was the scan configuration process?
    \item Did the real-time progress updates provide useful information?
    \item What are your thoughts on the visualization and presentation of scan results?
    \item How useful did you find the dashboard for getting an overview of your security posture?
\end{enumerate}

\section{Technical Aspects}

\begin{enumerate}
    \item Were there any technical limitations you encountered?
    \item Did you experience any performance issues?
    \item How adequate was the level of detail in the scan results?
    \item Were there any features you expected to find but didn't?
\end{enumerate}

\section{Practical Application}

\begin{enumerate}
    \item Can you see yourself using this system in your work?
    \item What use cases would you primarily use it for?
    \item What would make this system more valuable for your specific needs?
    \item Are there any concerns about deploying this in a production environment?
\end{enumerate}

\section{Future Enhancements}

\begin{enumerate}
    \item If you could add one feature to this system, what would it be?
    \item What integrations with other tools would be most valuable?
    \item How important is mobile access versus desktop-only?
    \item Would you be interested in advanced features like scheduled scanning or automated alerting?
\end{enumerate}

\clearpage

\chapter{Additional Supporting Material}
\label{app:additional}

\section{Installation Guide}

% [PLACEHOLDER: Installation instructions]

This section would contain detailed installation and deployment instructions including:

\begin{itemize}
    \item System prerequisites
    \item Backend setup (Python environment, dependencies)
    \item Frontend setup (Node.js, npm packages)
    \item Database configuration
    \item Security tool installation (Nmap, Masscan, Nikto)
    \item Configuration file templates
    \item Deployment options (development vs. production)
\end{itemize}

\section{API Documentation}

% [PLACEHOLDER: API endpoint documentation]

Comprehensive API documentation would include:

\begin{itemize}
    \item Endpoint specifications
    \item Request/response formats
    \item Authentication methods (when implemented)
    \item Error codes and handling
    \item Example requests using curl or similar tools
\end{itemize}

\section{Code Repository}

The complete source code for this project is available at:

% [PLACEHOLDER: Repository URL]
\texttt{https://github.com/[username]/threat-sentinel}

\section{Ethics Approval}

% [PLACEHOLDER: Ethics approval documentation]

This section would contain any ethics approval letters or documentation if required by the institution for conducting usability testing or other research activities involving human participants.

\section{Sample Scan Results}

% [PLACEHOLDER: Sample scan output examples]

This section would include representative examples of scan outputs for:

\begin{itemize}
    \item Host discovery scans
    \item Port scanning results
    \item Web vulnerability scan reports
    \item Exported data formats (JSON, CSV)
\end{itemize}
